---
title: "testflow"
output:
  github_document:
    md_extensions: -smart
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
}
```

![Version 0.8.2](https://img.shields.io/badge/version-0.8.2-blue.svg)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
![R >= 4.1.0](https://img.shields.io/badge/R-%3E%3D%204.1.0-blue.svg)

Current development release: `0.8.2`. First public release: `0.7.0`.

`testflow` helps choose, run, and report common statistical tests from the
study design. The package returns a consistent object with assumptions,
the recommended test, the null hypothesis, the result, an effect size when
appropriate, a plain-language report, and an optional plot.

The guiding idea is:

```r
testflow = design + assumptions + test + effect size + report + plot
```

## About

`testflow` is an R package for teaching and applied statistical workflows. It
keeps the statistical question organized around the study design, then combines
assumption checks, test selection, effect sizes, visual diagnostics, and
ready-to-use reporting in one object.

It is designed for analysts, students, and instructors who want a reproducible
path from a data question to a statistical test, a plot, and a concise written
interpretation.

## Installation

`testflow` is not on CRAN yet. Install the development version from GitHub:

```r
install.packages("remotes")
remotes::install_github("ielbadisy/testflow")
```

## Quick Start

```{r}
library(testflow)

cardio <- make_cardio_data()

res <- test_two_groups(sbp_3m ~ sex, data = cardio, plot = TRUE)
res
```

```{r}
report(res)
as_tibble(res)
```

```{r quick-start-plot, fig.width = 7, fig.height = 4.5}
plot(res)
```

Printed results use `cli` styling in interactive R sessions. GitHub strips
terminal colors, but the structure is the same in the console. To control
colors explicitly, use:

```r
options(testflow.cli_colors = FALSE)
options(testflow.cli_colors = TRUE)
```

## Complete Workflow

A typical analysis starts with the study question, lets `testflow` check the
assumptions, then uses the returned object for the report, tidy result, and
plot.

```{r complete-workflow}
paired <- test_paired(sbp_3m ~ sbp_baseline, data = cardio, plot = TRUE)

paired
report(paired)
as_tibble(paired)
```

```{r complete-workflow-plot, fig.width = 7, fig.height = 4.5}
plot(paired)
```

## Plot Gallery

Different designs produce different visual summaries. These examples use the
same teaching dataset so the function call, statistical result, and plot stay
together.

```{r plot-gallery-setup}
groups <- test_groups(sbp_3m ~ treatment, data = cardio, plot = TRUE)
repeated <- test_repeated(cardio, c(sbp_baseline, sbp_3m, sbp_6m), id = id, plot = TRUE)
correlation <- test_correlation(sbp_3m ~ age, data = cardio, plot = TRUE)
cor_matrix <- suppressWarnings(test_correlation_matrix(cardio, c(age, sbp_3m, ldl), plot = TRUE))
categorical <- test_categorical(treatment ~ controlled_3m, data = cardio, plot = TRUE)
outliers <- suppressWarnings(test_outliers(c(sbp_3m, ldl, crp), data = cardio, plot = TRUE))
```

### More Than Two Groups

```{r groups-plot, fig.width = 7, fig.height = 4.5}
plot(groups)
```

You can keep the default plot style while changing labels:

```{r custom-plot-labels, eval = FALSE}
plot(
  groups,
  title = "Systolic blood pressure by treatment",
  subtitle = "Automatically selected test and p-value are still available in the result object",
  caption = "Teaching dataset generated by make_cardio_data()"
)
```

### Repeated Measurements

```{r repeated-plot, fig.width = 7, fig.height = 4.5}
plot(repeated)
```

### Correlation

```{r correlation-plot, fig.width = 7, fig.height = 4.5}
plot(correlation)
```

### Correlation Matrix

Cell labels show the pairwise correlation coefficient (`r`).

```{r correlation-matrix-labeled-plot, fig.width = 6, fig.height = 5}
plot(cor_matrix)
```

### Categorical Association

```{r categorical-plot, fig.width = 7, fig.height = 4.5}
plot(categorical)
```

### Outlier Screening

For IQR screening, points below `Q1 - 1.5 x IQR` or above `Q3 + 1.5 x IQR`
are flagged. The dashed lines show those fences and red points are detected
outliers.

```{r outlier-plot, fig.width = 7, fig.height = 4.5}
plot(outliers)
```

## Summary Tables

`sumtab()` creates descriptive tables and can add automatically selected
p-values based on variable type and grouping structure.

```{r}
sumtab(~ age + sex + sbp_3m | treatment, cardio, p_value = TRUE)
```

## Sample Size Planning

`testflow` now includes planning functions for common sample-size problems.
The API is organized by endpoint family, design, and objective:

```{r, eval = FALSE}
sample_size(
  endpoint = "continuous",
  design = "parallel",
  objective = "superiority",
  delta = 5,
  sd = 10,
  alpha = 0.05,
  power = 0.90
)
```

Current helpers:

```{r, eval = FALSE}
sample_size_continuous()
sample_size_binary()
sample_size_survival()
sample_size_ordinal()
sample_size_adjust_dropout()
```

Supported planning settings:

| Endpoint | Supported design(s) | Supported objective(s) | Formula family |
|---|---|---|---|
| Continuous | parallel, paired, repeated(2 time points) | superiority, non-inferiority, equivalence | normal-approximation formulas for mean difference and paired differences |
| Binary | parallel, paired, repeated(2 time points) | superiority, non-inferiority, equivalence | risk-difference planning and discordant-pair planning |
| Survival | parallel | superiority, non-inferiority, equivalence | event-based proportional hazards planning |
| Ordinal | parallel | superiority | Noether approximation |

Example:

```{r, eval = FALSE}
paired_ss <- sample_size_continuous(
  design = "paired",
  objective = "superiority",
  delta = 5,
  sd_diff = 10,
  alpha = 0.05,
  power = 0.90
)

paired_ss
plot(paired_ss)
```

The formulas and reporting language are based on:

Julious, S. A. (2010). *Sample Sizes for Clinical Trials*. Chapman & Hall/CRC.

## Common Workflows

Use formulas when there is an outcome and a grouping, predictor, or repeated
measure structure. Use tidyselect-style column selection for multi-column
workflows such as repeated measures, correlation matrices, and outlier screens.

```{r eval = FALSE}
test_one_sample(cardio, sbp_3m, mu = 140)
test_two_groups(sbp_3m ~ sex, data = cardio)
test_paired(sbp_3m ~ sbp_baseline, data = cardio)
test_groups(sbp_3m ~ treatment, data = cardio)
test_factorial(sbp_3m ~ sex * treatment, data = cardio)
test_repeated(cardio, c(sbp_baseline, sbp_3m, sbp_6m), id = id)
test_categorical(treatment ~ controlled_3m, data = cardio)
test_paired_categorical(cardio, controlled_baseline, controlled_3m)
test_repeated_categorical(cardio, c(controlled_baseline, controlled_3m, controlled_6m))
test_proportion(cardio, controlled_3m, success = "yes", p = 0.5)
test_multinomial(cardio, treatment)
test_correlation(sbp_3m ~ age, data = cardio)
test_correlation_matrix(cardio, c(age, sbp_3m, ldl))
test_outliers(c(sbp_3m, ldl, crp), data = cardio)
sumtab(~ age + sex + sbp_3m | treatment, cardio, p_value = TRUE)
```

## What Is Implemented

| Study design | Main function | Tests considered |
|---|---|---|
| Summary table | `sumtab()` | Student t-test, Welch t-test, Wilcoxon rank-sum, ANOVA, Welch ANOVA, Kruskal-Wallis, chi-square, Fisher exact |
| One numeric sample | `test_one_sample()` | one-sample t-test, Wilcoxon signed-rank, sign test |
| Two independent groups | `test_two_groups()` | Student t-test, Welch t-test, Wilcoxon rank-sum |
| Paired numeric measurements | `test_paired()` | paired t-test, Wilcoxon signed-rank, sign test |
| More than two groups | `test_groups()` | one-way ANOVA, Welch ANOVA, Kruskal-Wallis, post hoc comparisons |
| Factorial numeric design | `test_factorial()` | factorial ANOVA with main effects and interactions |
| Repeated numeric measurements | `test_repeated()`, `test_repeated_long()` | repeated-measures ANOVA, Friedman test, paired post hoc tests |
| Two categorical variables | `test_categorical()` | chi-square independence test, Fisher exact test |
| Paired categorical measurements | `test_paired_categorical()` | McNemar test |
| Repeated categorical measurements | `test_repeated_categorical()` | Cochran Q test, pairwise McNemar tests |
| One proportion | `test_proportion()` | exact binomial test, one-sample proportion test |
| Multinomial categories | `test_multinomial()` | chi-square goodness-of-fit, pairwise binomial checks |
| Two numeric variables | `test_correlation()` | Pearson, Spearman, Kendall |
| Correlation matrix | `test_correlation_matrix()` | pairwise Pearson, Spearman, or Kendall correlations |
| Outlier screening | `test_outliers()` | IQR rule, Mahalanobis distance |

Effect-size formulas are documented in
`vignettes/effect-size-formulas.Rmd`.

## References

Bartlett, M. S. (1937). Properties of sufficiency and statistical tests.
*Proceedings of the Royal Society of London. Series A, Mathematical and
Physical Sciences, 160*(901), 268-282.

Clopper, C. J., & Pearson, E. S. (1934). The use of confidence or fiducial
limits illustrated in the case of the binomial. *Biometrika, 26*(4), 404-413.

Cochran, W. G. (1950). The comparison of percentages in matched samples.
*Biometrika, 37*(3/4), 256-266.

Cohen, J. (1988). *Statistical power analysis for the behavioral sciences*
(2nd ed.). Lawrence Erlbaum Associates.

Cramer, H. (1946). *Mathematical methods of statistics*. Princeton University
Press.

Dunn, O. J. (1964). Multiple comparisons using rank sums. *Technometrics,
6*(3), 241-252.

Fisher, R. A. (1925). *Statistical methods for research workers*. Oliver and
Boyd.

Friedman, M. (1937). The use of ranks to avoid the assumption of normality
implicit in the analysis of variance. *Journal of the American Statistical
Association, 32*(200), 675-701.

Kendall, M. G. (1938). A new measure of rank correlation. *Biometrika,
30*(1/2), 81-93.

Kruskal, W. H., & Wallis, W. A. (1952). Use of ranks in one-criterion variance
analysis. *Journal of the American Statistical Association, 47*(260), 583-621.

Levene, H. (1960). Robust tests for equality of variances. In I. Olkin,
S. G. Ghurye, W. Hoeffding, W. G. Madow, & H. B. Mann (Eds.),
*Contributions to probability and statistics: Essays in honor of Harold
Hotelling* (pp. 278-292). Stanford University Press.

Mahalanobis, P. C. (1936). On the generalized distance in statistics.
*Proceedings of the National Institute of Sciences of India, 2*(1), 49-55.

Mann, H. B., & Whitney, D. R. (1947). On a test of whether one of two random
variables is stochastically larger than the other. *The Annals of Mathematical
Statistics, 18*(1), 50-60.

McNemar, Q. (1947). Note on the sampling error of the difference between
correlated proportions or percentages. *Psychometrika, 12*(2), 153-157.

Pearson, K. (1895). Notes on regression and inheritance in the case of two
parents. *Proceedings of the Royal Society of London, 58*, 240-242.

Pearson, K. (1900). On the criterion that a given system of deviations from the
probable in the case of a correlated system of variables is such that it can be
reasonably supposed to have arisen from random sampling. *The London,
Edinburgh, and Dublin Philosophical Magazine and Journal of Science, 50*(302),
157-175.

Shapiro, S. S., & Wilk, M. B. (1965). An analysis of variance test for
normality (complete samples). *Biometrika, 52*(3/4), 591-611.

Spearman, C. (1904). The proof and measurement of association between two
things. *The American Journal of Psychology, 15*(1), 72-101.

Student. (1908). The probable error of a mean. *Biometrika, 6*(1), 1-25.

Tukey, J. W. (1949). Comparing individual means in the analysis of variance.
*Biometrics, 5*(2), 99-114.

Welch, B. L. (1947). The generalization of Student's problem when several
different population variances are involved. *Biometrika, 34*(1/2), 28-35.

Wilcoxon, F. (1945). Individual comparisons by ranking methods. *Biometrics
Bulletin, 1*(6), 80-83.
