---
title: "testflow"
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# testflow

`testflow` is organized around study design, not test names.
Instead of asking "which function runs a t-test?", testflow asks:
"what is the design of my study question?"

The core grammar is:

```r
testflow = test + interpret + plot
```

```{r}
library(testflow)

cardio <- make_cardio_data()

x <- cardio |>
  test_two_groups(sbp_3m ~ sex)

x
plot(x)
report(x)
as_tibble(x)
```

## Summary tables

Use `sumtab()` to build a descriptive table and, when requested, add
automatically selected p-values by variable type and group structure.

```{r}
sumtab(~ age + sex + sbp_3m | treatment, cardio, p_value = TRUE)
```

The public workflows are named for common study designs:

```{r eval = FALSE}
test_one_sample(cardio, sbp_3m, mu = 140)
test_two_groups(sbp_3m ~ sex, data = cardio)
test_paired(sbp_3m ~ sbp_baseline, data = cardio)
test_groups(sbp_3m ~ treatment, data = cardio)
test_factorial(sbp_3m ~ sex * treatment, data = cardio)
test_categorical(treatment ~ controlled_3m, data = cardio)
test_correlation(sbp_3m ~ age, data = cardio)
test_outliers(cardio, c(sbp_3m, ldl, crp))
sumtab(~ age + sex + sbp_3m | treatment, cardio, p_value = TRUE)
```

## Implemented workflows and tests

Each workflow returns a `testflow` object with the recommended test, H0, p-value,
confidence interval when available, appropriate effect size, report text, and plot.
For descriptive reporting, `sumtab()` builds a formula-driven summary table and
can add automatically selected p-values.

The exact formulas used for Cohen's d and the other reported effect-size
estimates are documented in `vignettes/effect-size-formulas.Rmd`.

| Workflow | Formula-oriented call | Tests considered |
|---|---|---|
| Summary table | `sumtab(~ age + sex + sbp_3m \| treatment, cardio, p_value = TRUE)` | Student t-test, Welch t-test, Wilcoxon rank-sum, ANOVA, Welch ANOVA, Kruskal-Wallis, chi-square, Fisher exact |
| One sample | `test_one_sample(cardio, sbp_3m, mu = 140)` | one-sample t-test, Wilcoxon signed-rank, sign test |
| Two independent groups | `test_two_groups(sbp_3m ~ sex, data = cardio)` | Student t-test, Welch t-test, Wilcoxon rank-sum |
| Paired measurements | `test_paired(sbp_3m ~ sbp_baseline, data = cardio)` | paired t-test, Wilcoxon signed-rank, sign test |
| More than two groups | `test_groups(sbp_3m ~ treatment, data = cardio)` | one-way ANOVA + Tukey, Welch ANOVA + Welch pairwise t-tests, Kruskal-Wallis + pairwise Wilcoxon |
| Factorial design | `test_factorial(sbp_3m ~ sex * treatment, data = cardio)` | factorial ANOVA with main effects and interactions |
| Repeated measurements | `test_repeated(cardio, c(sbp_baseline, sbp_3m, sbp_6m), id = id)` | repeated-measures ANOVA + paired t-tests, Friedman + paired Wilcoxon |
| Categorical association | `test_categorical(treatment ~ controlled_3m, data = cardio)` | chi-square independence test, Fisher exact test |
| Paired categorical | `test_paired_categorical(cardio, controlled_baseline, controlled_3m)` | McNemar test |
| Repeated categorical | `test_repeated_categorical(cardio, c(controlled_baseline, controlled_3m, controlled_6m))` | Cochran Q test + pairwise McNemar tests |
| One proportion | `test_proportion(cardio, controlled_3m, success = "yes", p = 0.5)` | exact binomial test, one-sample proportion test |
| Multinomial | `test_multinomial(cardio, treatment)` | chi-square goodness-of-fit, pairwise binomial checks |
| Correlation | `test_correlation(sbp_3m ~ age, data = cardio)` | Pearson, Spearman, Kendall |
| Correlation matrix | `test_correlation_matrix(cardio, c(age, sbp_3m, ldl))` | matrix of Pearson/Spearman/Kendall correlations |
| Outliers | `test_outliers(cardio, c(sbp_3m, ldl, crp))` | IQR outliers, Mahalanobis distance |

## References

- Fisher, R. A. (1925). \emph{Statistical Methods for Research Workers}.
- Gosset, W. S. (1908). The probable error of a mean.
- Welch, B. L. (1947). Generalization of Student's problem with unequal variances.
- Wilcoxon, F. (1945). Individual comparisons by ranking methods.
- Mann, H. B., & Whitney, D. R. (1947). On a test of whether one of two random variables is stochastically larger than the other.
- Levene, H. (1960). Robust tests for equality of variances.
- Kruskal, W. H., & Wallis, W. A. (1952). Use of ranks in one-criterion variance analysis.
- Tukey, J. W. (1949). Comparing individual means in the analysis of variance.
- Dunn, O. J. (1964). Multiple comparisons using rank sums.
- Friedman, M. (1937). The use of ranks to avoid the assumption of normality implicit in the analysis of variance.
- Cochran, W. G. (1950). The comparison of percentages in matched samples.
- McNemar, Q. (1947). Note on the sampling error of the difference between correlated proportions or percentages.
- Pearson, K. (1895, 1900).
- Spearman, C. (1904). The proof and measurement of association between two things.
- Kendall, M. G. (1938). A new measure of rank correlation.
- Cramer, H. (1946). \emph{Mathematical Methods of Statistics}.
- Clopper, C. J., & Pearson, E. S. (1934). The use of confidence or fiducial limits illustrated in the case of the binomial.
- Cohen, J. (1988). \emph{Statistical Power Analysis for the Behavioral Sciences}.
