---
title: "testflow"
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# testflow

`testflow` is organized around study design, not test names.
Instead of asking "which function runs a t-test?", testflow asks:
"what is the design of my clinical question?"

The core grammar is:

```r
testflow = test + interpret + plot
```

```{r eval = FALSE}
library(testflow)

cardio <- make_cardio_data()

x <- cardio |>
  test_two_groups(sbp_3m ~ sex)

x
plot(x)
report(x)
as_tibble(x)
```

The public workflows are named for common study designs:

```{r eval = FALSE}
test_one_sample(cardio, sbp_3m, mu = 140)
test_two_groups(sbp_3m ~ sex, data = cardio)
test_paired(cardio, sbp_baseline, sbp_3m)
test_groups(sbp_3m ~ treatment, data = cardio)
test_categorical(treatment ~ controlled_3m, data = cardio)
test_correlation(sbp_3m ~ age, data = cardio)
test_outliers(cardio, c(sbp_3m, ldl, crp))
```

## Implemented workflows and tests

Each workflow returns a `testflow` object with the recommended test, H0, p-value,
confidence interval when available, appropriate effect size, report text, and plot.

| Workflow | Formula-oriented call | Tests considered |
|---|---|---|
| One sample | `test_one_sample(cardio, sbp_3m, mu = 140)` | one-sample t-test, Wilcoxon signed-rank, sign test |
| Two independent groups | `test_two_groups(sbp_3m ~ sex, data = cardio)` | Student t-test, Welch t-test, Wilcoxon rank-sum |
| Paired measurements | `test_paired(cardio, sbp_baseline, sbp_3m)` | paired t-test, Wilcoxon signed-rank, sign test |
| More than two groups | `test_groups(sbp_3m ~ treatment, data = cardio)` | one-way ANOVA, Welch ANOVA, Kruskal-Wallis, post-hoc comparisons |
| Factorial design | `test_factorial(cardio, sbp_3m, c(sex, treatment))` | factorial ANOVA with main effects and interactions |
| Repeated measurements | `test_repeated(cardio, c(sbp_baseline, sbp_3m, sbp_6m), id = id)` | repeated-measures workflow, Friedman alternative |
| Categorical association | `test_categorical(treatment ~ controlled_3m, data = cardio)` | chi-square independence test, Fisher exact test |
| Paired categorical | `test_paired_categorical(cardio, controlled_baseline, controlled_3m)` | McNemar test |
| Repeated categorical | `test_repeated_categorical(cardio, c(controlled_baseline, controlled_3m, controlled_6m))` | Cochran Q-style repeated categorical workflow |
| One proportion | `test_proportion(cardio, controlled_3m, success = "yes", p = 0.5)` | exact binomial test, one-sample proportion test |
| Multinomial | `test_multinomial(cardio, treatment)` | chi-square goodness-of-fit, pairwise binomial checks |
| Correlation | `test_correlation(sbp_3m ~ age, data = cardio)` | Pearson, Spearman, Kendall |
| Correlation matrix | `test_correlation_matrix(cardio, c(age, sbp_3m, ldl))` | matrix of Pearson/Spearman/Kendall correlations |
| Outliers | `test_outliers(cardio, c(sbp_3m, ldl, crp))` | IQR outliers, Mahalanobis distance |
