---
title: "Sample size: complete API reference and formulas"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Sample size: complete API reference and formulas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE)
```

```{r setup}
library(testflow)
```

## Purpose

The "Sample size planning" vignette introduces the `sample_size()` API and
its main use cases. This vignette is a complete reference: every
`design`/`objective`/`method` combination each planning function accepts,
paired with the exact formula it evaluates, and a runnable example. Use it
as a lookup table when you already know which planning function you need
and want to see (or check) precisely what it computes.

Notation used throughout, matching the function documentation:

- \(\alpha\): type I error rate; \(1-\beta\): power.
- \(z_{1-\alpha}\), \(z_{1-\alpha/2}\): one- and two-sided standard normal
  quantiles; \(z_{1-\beta}\): power quantile.
- \(r = n_B/n_A\): allocation ratio (`allocation`), with \(n_B = rn_A\).
- Dropout adjustment is always \(n_{recruit} = \lceil n_{eval}/(1-d)\rceil\).

Superiority tests use two-sided \(\alpha\); non-inferiority and equivalence
use one-sided \(\alpha\) (pass `alpha = 0.025` for a one-sided-0.025 design,
or keep `alpha = 0.05` for a one-sided-0.05 design). `testflow` does not
rescale `alpha` for you.

## `sample_size_continuous()`

```{r, eval = FALSE}
sample_size_continuous(
  design = c("parallel", "paired", "repeated"),
  objective = c("superiority", "noninferiority", "equivalence"),
  delta = NULL, sd = NULL, sd_diff = NULL, expected_difference = 0,
  margin = NULL, alpha = 0.05, power = 0.90, allocation = 1, dropout = 0,
  n_time = 2, correlation = 0.5
)
```

`delta` doubles as the superiority effect size *or* the non-inferiority/
equivalence margin, depending on `objective`; `expected_difference` is the
planned/observed difference \(\Delta\) used only for non-inferiority and
equivalence.

### `design = "parallel"`

#### `objective = "superiority"`

\[
n_A = \frac{(r+1)\sigma^2(z_{1-\beta}+z_{1-\alpha/2})^2}{r\Delta^2},
\qquad n_B = rn_A
\]

```{r}
sample_size_continuous(
  design = "parallel", objective = "superiority",
  delta = 5, sd = 10, allocation = 1, alpha = 0.05, power = 0.90
)
```

#### `objective = "noninferiority"`

With \(d_{NI} = \Delta + \delta\) (`delta` is the margin \(\delta\)),
requiring \(d_{NI} > 0\):

\[
n_A = \frac{(r+1)\sigma^2(z_{1-\beta}+z_{1-\alpha})^2}{rd_{NI}^2}
\]

```{r}
sample_size_continuous(
  design = "parallel", objective = "noninferiority",
  delta = 2, expected_difference = 0.5, sd = 10, alpha = 0.025, power = 0.90
)
```

#### `objective = "equivalence"`

General case, with \(d_{EQ} = \delta - |\Delta|\), requiring \(d_{EQ} > 0\):

\[
n_A \approx \frac{(r+1)\sigma^2(z_{1-\beta}+z_{1-\alpha})^2}{rd_{EQ}^2}
\]

```{r}
sample_size_continuous(
  design = "parallel", objective = "equivalence",
  delta = 4, expected_difference = 1, sd = 10, alpha = 0.05, power = 0.90
)
```

Special case \(\Delta = 0\) (`testflow` switches formulas automatically):

\[
n_A = \frac{(r+1)\sigma^2(z_{1-\beta/2}+z_{1-\alpha})^2}{r\delta^2}
\]

```{r}
sample_size_continuous(
  design = "parallel", objective = "equivalence",
  delta = 4, expected_difference = 0, sd = 10, alpha = 0.05, power = 0.90
)
```

### `design = "paired"`

Paired designs use the same three forms with the paired-difference standard
deviation `sd_diff` in place of \(\sigma\), and no allocation ratio (a
paired analysis is a one-sample test on the differences):

\[
n_{pairs} = \frac{(z_{1-\alpha/2}+z_{1-\beta})^2\sigma_{diff}^2}{\Delta^2}
\qquad\text{(superiority)}
\]

```{r}
sample_size_continuous(
  design = "paired", objective = "superiority",
  delta = 5, sd_diff = 10, alpha = 0.05, power = 0.90
)

sample_size_continuous(
  design = "paired", objective = "noninferiority",
  delta = 2, expected_difference = 0.5, sd_diff = 10, alpha = 0.025, power = 0.90
)

sample_size_continuous(
  design = "paired", objective = "equivalence",
  delta = 4, expected_difference = 0, sd_diff = 10, alpha = 0.05, power = 0.90
)
```

If `sd_diff` is omitted, `sd` is used in its place.

### `design = "repeated"`

With `n_time = 2` this is identical to `design = "paired"`. With
`n_time > 2`, `sd_diff` (or `sd`) is rescaled to an effective standard
deviation under a compound-symmetry working correlation before being
plugged into the same paired formulas above:

\[
\sigma_{eff} = \sigma_{diff}\sqrt{\frac{1+(m-1)\rho}{m}}
\]

where \(m\) is `n_time` and \(\rho\) is `correlation`. This design-effect
step is a `testflow` extension for planning an average treatment-effect
test across exchangeably correlated repeated measurements; it is not itself
a closed-form textbook formula.

```{r}
sample_size_continuous(
  design = "repeated", n_time = 4, correlation = 0.5,
  objective = "superiority", delta = 5, sd_diff = 10, alpha = 0.05, power = 0.90
)

sample_size_continuous(
  design = "repeated", n_time = 4, correlation = 0.3,
  objective = "noninferiority", delta = 2, expected_difference = 0.5,
  sd_diff = 10, alpha = 0.025, power = 0.90
)

sample_size_continuous(
  design = "repeated", n_time = 4, correlation = 0.3,
  objective = "equivalence", delta = 4, expected_difference = 0,
  sd_diff = 10, alpha = 0.05, power = 0.90
)
```

## `sample_size_binary()`

```{r, eval = FALSE}
sample_size_binary(
  design = c("parallel", "paired", "repeated"),
  objective = c("superiority", "noninferiority", "equivalence"),
  p1, p2, margin = NULL, method = c("pooled", "anticipated"),
  discordant_or = NULL, discordance_rate = NULL, p10 = NULL, p01 = NULL,
  alpha = 0.05, power = 0.90, allocation = 1, dropout = 0, n_time = 2
)
```

### `design = "parallel"`

#### `objective = "superiority"`, `method = "pooled"`

With pooled proportion \(\bar p = (p_A+rp_B)/(1+r)\):

\[
n_A = \frac{\left[z_{1-\alpha/2}\sqrt{(1+1/r)\bar p(1-\bar p)}
  + z_{1-\beta}\sqrt{p_A(1-p_A)+p_B(1-p_B)/r}\right]^2}{(p_A-p_B)^2}
\]

```{r}
sample_size_binary(
  design = "parallel", objective = "superiority",
  p1 = 0.4, p2 = 0.25, method = "pooled", alpha = 0.05, power = 0.90
)
```

#### `objective = "superiority"`, `method = "anticipated"`

\[
n_A \approx \frac{(z_{1-\alpha/2}+z_{1-\beta})^2
  \left[p_A(1-p_A)+p_B(1-p_B)/r\right]}{(p_A-p_B)^2}
\]

```{r}
sample_size_binary(
  design = "parallel", objective = "superiority",
  p1 = 0.4, p2 = 0.25, method = "anticipated", alpha = 0.05, power = 0.90
)
```

#### `objective = "noninferiority"`

With \(d_{NI} = (p_A-p_B)+\delta\) (`margin` is \(\delta\)):

\[
n_A \approx \frac{(z_{1-\alpha}+z_{1-\beta})^2
  \left[p_A(1-p_A)+p_B(1-p_B)/r\right]}{d_{NI}^2}
\]

```{r}
sample_size_binary(
  design = "parallel", objective = "noninferiority",
  p1 = 0.5, p2 = 0.45, margin = 0.1, alpha = 0.025, power = 0.90
)
```

#### `objective = "equivalence"`

General case, with \(d_{EQ} = \delta - |p_A-p_B|\):

\[
n_A \approx \frac{(z_{1-\alpha}+z_{1-\beta})^2
  \left[p_A(1-p_A)+p_B(1-p_B)/r\right]}{d_{EQ}^2}
\]

```{r}
sample_size_binary(
  design = "parallel", objective = "equivalence",
  p1 = 0.42, p2 = 0.4, margin = 0.15, alpha = 0.05, power = 0.90
)
```

Special case \(p_A=p_B=p\), which still scales with the allocation ratio
`r` through an \((r+1)/r\) factor:

\[
n_A = \frac{(r+1)(z_{1-\beta/2}+z_{1-\alpha})^2p(1-p)}{r\delta^2}
\]

```{r}
sample_size_binary(
  design = "parallel", objective = "equivalence",
  p1 = 0.3, p2 = 0.3, margin = 0.15, allocation = 2, alpha = 0.05, power = 0.90
)
```

### `design = "paired"` / `design = "repeated"` (`n_time = 2`)

Discordant-pairs (McNemar-type) planning, entered either directly via
`discordant_or`/`discordance_rate`, or via the discordant-cell probabilities
`p10` (A-only) and `p01` (B-only), from which
\(OR_D = p_{10}/p_{01}\) and \(\lambda_D = p_{10}+p_{01}\) are derived:

\[
n_{total} = \left\lceil
\frac{(z_{1-\alpha/2}+z_{1-\beta})^2(OR_D+1)^2}{(OR_D-1)^2\lambda_D}
\right\rceil
\]

```{r}
sample_size_binary(
  design = "paired", objective = "superiority",
  discordant_or = 2, discordance_rate = 0.3, alpha = 0.05, power = 0.90
)

sample_size_binary(
  design = "paired", objective = "superiority",
  p10 = 0.2, p01 = 0.1, alpha = 0.05, power = 0.90
)

sample_size_binary(
  design = "repeated", n_time = 2, objective = "superiority",
  p10 = 0.2, p01 = 0.1, alpha = 0.05, power = 0.90
)
```

`design = "repeated"` for binary endpoints only supports `n_time = 2` (it is
routed to the same discordant-pairs formula as `design = "paired"`);
`n_time > 2` is not implemented for binary endpoints.

## `sample_size_survival()`

```{r, eval = FALSE}
sample_size_survival(
  design = c("parallel"),
  objective = c("superiority", "noninferiority", "equivalence"),
  hr, margin_hr = NULL, lower = NULL, upper = NULL,
  survival_a = NULL, survival_b = NULL,
  alpha = 0.05, power = 0.90, allocation = 1, dropout = 0,
  method = c("exponential", "ph_only")
)
```

Only `design = "parallel"` with equal allocation is implemented.

### `objective = "superiority"`, `method = "exponential"`

Required total events, assuming an exponential hazard:

\[
D_{total} = \frac{4(z_{1-\alpha/2}+z_{1-\beta})^2}{[\log(HR)]^2}
\]

```{r}
sample_size_survival(
  hr = 0.7, method = "exponential", alpha = 0.05, power = 0.90
)
```

### `objective = "superiority"`, `method = "ph_only"`

Proportional-hazards-only approximation:

\[
D_{total} = \frac{2(HR+1)^2(z_{1-\alpha/2}+z_{1-\beta})^2}{(HR-1)^2}
\]

```{r}
sample_size_survival(
  hr = 0.7, method = "ph_only", alpha = 0.05, power = 0.90
)
```

### Converting events to a total sample size

Supplying `survival_a`/`survival_b` (planned survival probabilities at the
analysis horizon) converts required events to a total sample size, under
equal allocation:

\[
N_{total} = \frac{2D_{total}}{(1-S_A(t))+(1-S_B(t))}
\]

which follows from setting expected events
\(\tfrac{N}{2}(1-S_A(t)) + \tfrac{N}{2}(1-S_B(t))\) equal to \(D_{total}\).
Without `survival_a`/`survival_b`, `n` is the required event count itself.

```{r}
sample_size_survival(
  hr = 0.7, survival_a = 0.8, survival_b = 0.7, alpha = 0.05, power = 0.90
)
```

### `objective = "noninferiority"`

On the log-hazard-ratio scale, with \(d_{NI} = \log(HR_M) - \log(HR)\)
(`margin_hr` is \(HR_M\)), requiring \(d_{NI} > 0\):

\[
D_{total} \approx \frac{4(z_{1-\alpha}+z_{1-\beta})^2}{d_{NI}^2}
\]

```{r}
sample_size_survival(
  hr = 0.85, objective = "noninferiority", margin_hr = 1.25,
  survival_a = 0.75, survival_b = 0.75, alpha = 0.025, power = 0.90
)
```

### `objective = "equivalence"`

With log-hazard-ratio bounds `lower`/`upper` (\(\theta_L,\theta_U\)), and
\(d_{EQ} = \min(\theta-\theta_L,\ \theta_U-\theta)\) where
\(\theta=\log(HR)\), requiring \(d_{EQ} > 0\):

\[
D_{total} \approx \frac{4(z_{1-\alpha}+z_{1-\beta})^2}{d_{EQ}^2}
\]

```{r}
sample_size_survival(
  hr = 1.0, objective = "equivalence", lower = 0.8, upper = 1.25,
  survival_a = 0.75, survival_b = 0.75, alpha = 0.05, power = 0.90
)
```

## `sample_size_ordinal()`

```{r, eval = FALSE}
sample_size_ordinal(
  design = c("parallel"), objective = c("superiority"),
  p_superiority, alpha = 0.05, power = 0.90, dropout = 0
)
```

Only `design = "parallel"` and `objective = "superiority"` are implemented.
Noether's method for two independent groups, with
\(P = P(A>B) + \tfrac12P(A=B)\) (`p_superiority`, must exceed 0.5):

\[
n_{per\ group} = \frac{(z_{1-\alpha/2}+z_{1-\beta})^2}{6(P-0.5)^2}
\]

```{r}
sample_size_ordinal(p_superiority = 0.65, alpha = 0.05, power = 0.90)
```

## `sample_size()` dispatcher

`sample_size()` routes to the four functions above by `endpoint`:

```{r}
sample_size(
  endpoint = "binary", design = "parallel", objective = "noninferiority",
  p1 = 0.5, p2 = 0.45, margin = 0.1, alpha = 0.025, power = 0.90
)
```

## `sample_size_adjust_dropout()`

Exposes the dropout adjustment directly, independent of the planning
functions above:

\[
n_{recruit} = \left\lceil \frac{n_{eval}}{1-d} \right\rceil
\]

```{r}
sample_size_adjust_dropout(100, dropout = 0.15)
```

## Methods on `sample_size` objects

Every call above returns a `sample_size` object with the following methods:

```{r}
x <- sample_size_continuous(
  design = "paired", objective = "superiority",
  delta = 5, sd_diff = 10, alpha = 0.05, power = 0.90
)

x                       # print.sample_size(): formatted console report
summary(x)              # summary.sample_size(): compact summary list
report(x)               # report.sample_size(): report-ready sentence
as_tibble(x)            # as_tibble.sample_size(): one-row tidy summary
plot(x, type = "summary") # raw vs. dropout-adjusted n bar chart
plot(x, type = "curve")   # power curve, when curve data is available
plot(x, type = "both")    # both plots stacked (requires the patchwork package)
```

`curve_data` (and therefore `plot(x, type = "curve"/"both")`) is currently
only populated for `sample_size_continuous(design = "paired"/"repeated",
objective = "superiority")`; other endpoint/design/objective combinations
return a `NULL` curve and fall back to the summary bar chart.

None of these objects carry a `formula` or `reference` field. The formula
for each specific call is documented on its function's help page
(`?sample_size_continuous`, `?sample_size_binary`, `?sample_size_survival`,
`?sample_size_ordinal`) and reproduced above.

## Reference

Julious SA. *Sample Sizes for Clinical Trials*. Chapman & Hall/CRC; 2010.
