---
title: "Statistical test workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical test workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

`testflow` provides statistical testing workflows organized by clinical study design.

## One numerical variable

```{r eval = FALSE}
test_one_sample(cardio, sbp_3m, mu = 140)
```

## Two independent groups

```{r eval = FALSE}
test_two_groups(sbp_3m ~ sex, data = cardio)
```

## Paired measurements

```{r eval = FALSE}
test_paired(cardio, sbp_baseline, sbp_3m)
```

## More than two groups

```{r eval = FALSE}
test_groups(sbp_3m ~ treatment, data = cardio)
```

## Repeated measurements

```{r eval = FALSE}
test_repeated(cardio, c(sbp_baseline, sbp_3m, sbp_6m), id = id)
```

## Categorical outcomes

```{r eval = FALSE}
test_categorical(treatment ~ controlled_3m, data = cardio)
```

## Correlation

```{r eval = FALSE}
test_correlation(sbp_3m ~ age, data = cardio)
```

## Outliers

```{r eval = FALSE}
test_outliers(cardio, c(sbp_3m, ldl, crp))
```

## Reporting and plotting

Every workflow returns a `testflow` object. Use `report(x)`, `plot(x)`, and `as_tibble(x)`.
