---
title: "testflow: complete pedagogical reference"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{testflow: complete pedagogical reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE, fig.width = 6, fig.height = 4)
```

```{r setup}
library(testflow)
cardio <- make_cardio_data()
```

# Introduction

This vignette is the complete pedagogical reference for `testflow`: every
exported workflow function, organized by study design, alongside the exact
mathematical formula each one evaluates. It supersedes and consolidates the
previous `effect-size-formulas` and `statistical-test-workflows` vignettes,
and folds in the sample-size planning module so that every formula in the
package - hypothesis testing, effect sizes, and sample-size planning - lives
in one place.

Every `test_*()` function returns a `testflow` object: a list containing the
cleaned data, descriptive statistics, assumption checks, the recommended
test with its rationale, the primary test result (with its null hypothesis),
alternative test results for comparison, an effect size, an optional
`ggplot`, the original call, and a plain-language interpretation. Four
methods work on every object:

```{r}
x <- test_two_groups(sbp_3m ~ sex, data = cardio)
x                # print.testflow(): formatted console report
report(x)        # report.testflow(): report-ready sentence
as_tibble(x)      # as_tibble.testflow(): one-row tidy summary
# plot(x)         # plot.testflow(): the workflow's ggplot
```

Notation used throughout: \(n\) is a sample size, \(\bar x\) and \(s_x\) are
a sample mean and standard deviation, \(z_p\) and \(t_{df,p}\) are the
\(p\)-th quantiles of the standard normal and \(t\) distributions, and
\(\chi^2_{df}\), \(F_{df_1,df_2}\) name a chi-square and \(F\) distribution
with the given degrees of freedom.

## How test selection works

Most `test_*()` functions run assumption checks first and use them to pick
among two or three candidate tests (the *recommended* test), while still
computing and reporting every candidate under `alternative_tests` so you can
compare. The decision rules (`R/recommend.R`) are simple and deliberately
transparent:

| Workflow | Normal? & equal variance? | Normal only | Neither |
|---|---|---|---|
| One sample | One-sample t-test | - | Wilcoxon signed-rank |
| Two groups | Student t-test | Welch t-test | Wilcoxon rank-sum |
| Paired | Paired t-test | - | Wilcoxon signed-rank |
| 3+ groups | One-way ANOVA | Welch ANOVA | Kruskal-Wallis |

Normality is checked with Shapiro-Wilk (per group, where applicable);
variance homogeneity with Levene's test. `test_categorical()` instead
compares chi-square vs. Fisher's exact test based on expected cell counts
(`fisher_threshold`, default 5); `test_correlation(method = "auto")`
compares Pearson vs. Spearman based on normality and outlier screens; the
numeric repeated-measures workflow compares repeated-measures ANOVA vs.
Friedman based on within-time normality.

# Descriptive statistics

`sumtab()` and the internal `descriptives_numeric()`/`descriptives_categorical()`
helpers (populating every `testflow` object's `$descriptives` field) report,
for numeric variables:

\[
n,\quad \bar x = \frac{1}{n}\sum_i x_i,\quad
s_x = \sqrt{\frac{1}{n-1}\sum_i (x_i-\bar x)^2},\quad
\text{median},\quad IQR = Q_3-Q_1,\quad \min,\ \max
\]

and, for categorical variables, the count and percentage of each level:

```{r}
sumtab(~ age + sbp_3m + ldl | treatment, data = cardio)
```

# Assumption checks

`testflow` runs a small set of standardized assumption checks and reports
them in every object's `$assumptions` field, each with a `status`
(`acceptable`/`not acceptable`/`warning`/`not applicable`/`assumed`/`not
checked`) driving the recommended-test logic above.

| Check | Method | Statistic |
|---|---|---|
| Normality | Shapiro-Wilk (`n <= 5000`) | \(W\) |
| Symmetry (one-sample/paired) | Cabilio-Masaro | \(z=\sqrt n(\bar x-\text{median})/(s_x\sqrt{\pi/2-1})\) |
| Variance homogeneity (2 groups) | `var.test()` (F-test) | \(F=s_1^2/s_2^2\) |
| Variance homogeneity (3+ groups) | Levene's test | \(F\) on absolute deviations from group medians |
| Variance homogeneity (3+ groups) | Bartlett's test | \(\chi^2\) on group variances |
| Sphericity (3+ repeated measures) | Mauchly's test | \(W\) |
| Linearity (correlation) | Quadratic-term F-test (RESET, power 2) | \(F\) comparing `y ~ x` to `y ~ x + I(x^2)` |
| Monotonicity (correlation) | Spearman correlation of \(x,y\) | \(\rho\) |
| Expected cell counts (categorical) | Pearson chi-square expected counts | \(\min(E_{ij}) \geq\) threshold (default 5) |
| Multicollinearity (regression) | Variance inflation factor | \(VIF_j = 1/(1-R_j^2)\), flagged at \(\geq 5\) |
| Influential points (logistic regression) | Cook's distance | flagged at \(> 4/n\) |
| Proportional hazards (Cox) | Schoenfeld residual test (Grambsch-Therneau) | \(\chi^2\) |
| Outliers (screen) | IQR rule | \(x < Q_1-1.5\,IQR\) or \(x > Q_3+1.5\,IQR\) |

Independence of observations is *assumed* from the study design in every
workflow (`testflow` has no way to test it from the data alone) and is
always listed explicitly rather than silently taken for granted.

# One numerical variable

`test_one_sample()` compares a numeric sample against a fixed reference
value \(\mu_0\). The recommended test follows normality of the sample:

\[
t = \frac{\bar x - \mu_0}{s_x/\sqrt n} \sim t_{n-1}
\qquad\text{(One-sample t-test, Gosset 1908)}
\]

with the Wilcoxon signed-rank test as the non-parametric alternative, and a
sign test always reported alongside for reference. The reported effect size
is Cohen's \(d\):

\[
d = \frac{\bar x - \mu_0}{s_x}
\]

```{r}
test_one_sample(cardio, sbp_3m, mu = 140)
```

# Two independent groups

`test_two_groups()` compares a numeric outcome between two independent
groups, selecting among the Student t-test, Welch t-test, and Wilcoxon
rank-sum test per the table above:

\[
t = \frac{\bar x_1-\bar x_2}{s_p\sqrt{1/n_1+1/n_2}} \sim t_{n_1+n_2-2},
\qquad
s_p = \sqrt{\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2}}
\qquad\text{(Student/Gosset 1908)}
\]

Welch's t-test (1947) instead uses \(s_1^2/n_1+s_2^2/n_2\) as the variance
estimate with a Satterthwaite-approximated denominator degrees of freedom -
appropriate when variances are unequal. The Wilcoxon rank-sum (Mann-Whitney)
test compares rank sums rather than means. Effect size is Cohen's \(d\)
(t-test branches) or the rank-biserial correlation \(r_{rb}=1-2W/(n_1n_2)\)
(Wilcoxon branch), both with the sign following "first group minus second
group."

```{r}
test_two_groups(sbp_3m ~ sex, data = cardio)
```

# Paired measurements

`test_paired()` compares two measurements on the same subjects via their
paired differences \(d_i = \text{after}_i - \text{before}_i\):

\[
t = \frac{\bar d}{s_d/\sqrt n} \sim t_{n-1}
\qquad\text{(Paired t-test)}
\]

with the Wilcoxon signed-rank test on the differences as the non-parametric
alternative (selected when the differences fail the normality check), and a
sign test reported for reference. Effect size is Cohen's \(d_z=\bar d/s_d\).

```{r}
test_paired(sbp_3m ~ sbp_baseline, data = cardio)
```

# More than two groups

`test_groups()` extends the two-group comparison to \(k\geq 3\) groups:

\[
F = \frac{SS_{between}/(k-1)}{SS_{within}/(n-k)} \sim F_{k-1,\,n-k}
\qquad\text{(One-way ANOVA, Fisher 1925)}
\]

Welch's ANOVA relaxes the equal-variance assumption (analogous to Welch's
t-test); the Kruskal-Wallis test is the rank-based, distribution-free
alternative:

\[
H = \frac{12}{n(n+1)}\sum_{j=1}^k \frac{R_j^2}{n_j} - 3(n+1) \sim \chi^2_{k-1}
\qquad\text{(Kruskal-Wallis, 1952)}
\]

Effect size is eta squared \(\eta^2 = SS_{between}/SS_{total}\) for the
ANOVA branches, or Kruskal's epsilon squared
\(\epsilon^2 = (H-k+1)/(n-k)\) for the Kruskal-Wallis branch.

```{r}
test_groups(sbp_3m ~ treatment, data = cardio)
```

## Post-hoc comparisons

When `posthoc = TRUE` (the default), pairwise comparisons follow the
selected omnibus test: Tukey's HSD (1949) after one-way ANOVA, BH-adjusted
pairwise Welch t-tests after Welch ANOVA, and BH-adjusted pairwise Wilcoxon
tests (Dunn-style multiple comparisons, 1964) after Kruskal-Wallis.

# Factorial designs

`test_factorial()` fits a two- (or more-) way ANOVA with `outcome ~ f1 * f2`
and reports sums of squares of the chosen `type`:

- **Type I** (`stats::anova()`): sequential SS, order-dependent for
  unbalanced designs.
- **Type II**/**Type III** (`car::Anova()`): each term's SS adjusted for
  every other term (Type II) or additionally for the intercept under
  sum-to-zero contrasts (Type III, `testflow` sets `contr.sum` automatically
  since Type III under the default treatment contrasts tests a different,
  usually meaningless, hypothesis).

Effect size is eta squared per term, computed from whichever SS table
(Type I/II/III) was actually used for the reported test - not always
recomputed as Type I - so the effect size matches the p-value. Residual
normality (Shapiro-Wilk) and variance homogeneity (Levene, on the first
factor) are checked as assumptions.

```{r}
test_factorial(sbp_3m ~ sex * treatment, data = cardio)
```

# Repeated measurements (numeric)

`test_repeated()` (wide data) and `test_repeated_long()` (long data) compare
3 or more repeated numeric measurements on the same subjects, choosing
between repeated-measures ANOVA and the Friedman test based on within-time
normality:

\[
F = \frac{MS_{time}}{MS_{time \times subject}} \sim F_{k-1,\,(k-1)(n-1)}
\qquad\text{(Repeated-measures ANOVA, via `Error(id/time)`)}
\]

\[
\chi_F^2 = \frac{12}{nk(k+1)}\sum_{j=1}^k R_j^2 - 3n(k+1) \sim \chi^2_{k-1}
\qquad\text{(Friedman, 1937)}
\]

Sphericity (Mauchly's test) is checked for the ANOVA branch. Effect size is
\(\eta^2_{time}=SS_{time}/SS_{total}\) (ANOVA branch) or Kendall's
\(W=\chi_F^2/(n(k-1))\) (Friedman branch). Post-hoc comparisons are
BH-adjusted pairwise paired t-tests (ANOVA branch) or pairwise paired
Wilcoxon tests (Friedman branch).

```{r}
test_repeated(cardio, c(sbp_baseline, sbp_3m, sbp_6m), id = id)
```

# Categorical outcomes

## Association between two categorical variables

`test_categorical()` tests independence in a contingency table, selecting
chi-square or Fisher's exact test based on expected cell counts:

\[
\chi^2 = \sum_{i,j}\frac{(O_{ij}-E_{ij})^2}{E_{ij}} \sim \chi^2_{(r-1)(c-1)}
\qquad\text{(Pearson 1900)}
\]

Fisher's exact test (1922) is used instead whenever any expected count
falls below `fisher_threshold` (default 5). Effect size is Cramer's V:

\[
V = \sqrt{\frac{\chi^2}{n(\min(r,c)-1)}}
\]

```{r}
test_categorical(treatment ~ controlled_3m, data = cardio)
```

## One-sample proportion

`test_proportion()` tests a single proportion against a reference value
\(p_0\), using the exact binomial test (Clopper-Pearson-based) when expected
successes or failures fall below 5, or the standard proportion test
otherwise:

\[
z = \frac{\hat p - p_0}{\sqrt{p_0(1-p_0)/n}}
\qquad\text{(one-sample proportion test, normal approximation)}
\]

```{r}
test_proportion(cardio, controlled_3m, success = "yes", p = 0.5)
```

## Multinomial goodness of fit

`test_multinomial()` compares observed category counts to expected
probabilities \(p_1,\ldots,p_k\) (equal by default):

\[
\chi^2 = \sum_{i=1}^k \frac{(O_i-E_i)^2}{E_i} \sim \chi^2_{k-1}
\]

BH-adjusted per-category binomial tests are reported alongside as post-hoc
comparisons.

```{r}
test_multinomial(cardio, treatment)
```

## Paired categorical measurements

`test_paired_categorical()` compares a binary categorical outcome measured
twice on the same subjects via McNemar's test on the discordant cells
\(b\) (before-only) and \(c\) (after-only) of the 2x2 table:

\[
\chi^2 = \frac{(b-c)^2}{b+c} \sim \chi^2_1
\qquad\text{(McNemar, 1947)}
\]

Effect size is the number of discordant pairs \(b+c\).

```{r}
test_paired_categorical(cardio, controlled_baseline, controlled_3m)
```

## Repeated categorical measurements (3+ time points)

`test_repeated_categorical()` extends McNemar's test to \(k\geq 3\) repeated
binary measurements via Cochran's Q test:

\[
Q = \frac{(k-1)\left[k\sum_j C_j^2 - \left(\sum_j C_j\right)^2\right]}
{k\sum_i R_i - \sum_i R_i^2} \sim \chi^2_{k-1}
\qquad\text{(Cochran, 1950)}
\]

where \(C_j\) is the column (time) total and \(R_i\) the row (subject)
total. Effect size is the Cochran-Q analogue of Kendall's \(W\),
\(W=Q/(n(k-1))\); pairwise McNemar tests (BH-adjusted) are the post-hoc
comparisons.

```{r}
test_repeated_categorical(cardio, c(controlled_baseline, controlled_3m, controlled_6m))
```

# Correlation

`test_correlation()` tests association between two numeric variables, with
`method = "auto"` choosing Pearson when both variables pass the normality
and outlier screens, and Spearman otherwise:

\[
r = \frac{\sum_i(x_i-\bar x)(y_i-\bar y)}{\sqrt{\sum_i(x_i-\bar x)^2}\sqrt{\sum_i(y_i-\bar y)^2}}
\qquad\text{(Pearson, 1895)}
\]

\[
\rho = 1-\frac{6\sum_i d_i^2}{n(n^2-1)}, \quad d_i = \text{rank}(x_i)-\text{rank}(y_i)
\qquad\text{(Spearman, 1904)}
\]

Kendall's \(\tau\) (1938), based on the number of concordant vs. discordant
pairs, is always computed and reported in `alternative_tests`. Each of the
three is a candidate effect size, reported as the estimate of the selected
method. Linearity (for Pearson) and monotonicity (for Spearman/Kendall) are
checked as assumptions.

```{r}
test_correlation(sbp_3m ~ age, data = cardio)
```

`test_correlation_matrix()` extends this to all pairwise correlations among
3+ numeric variables, using pairwise-complete observations per pair and
Benjamini-Hochberg-adjusted p-values across all pairs.

```{r}
test_correlation_matrix(cardio, c(age, sbp_baseline, ldl, crp), method = "spearman")
```

# Regression

## Linear regression

`test_linear_regression()` fits ordinary least squares:

\[
y_i = \beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}+\varepsilon_i
\]

and reports the overall model F-test as the primary result:

\[
F=\frac{(SS_{total}-SS_{res})/p}{SS_{res}/(n-p-1)} \sim F_{p,\,n-p-1}
\]

\[
R^2=1-\frac{SS_{res}}{SS_{total}}, \qquad
R^2_{adj}=1-(1-R^2)\frac{n-1}{n-p-1}
\]

Per-term coefficients with confidence intervals are in
`alternative_tests$coefficients`. Residual normality (Shapiro-Wilk),
homoscedasticity (Breusch-Pagan score test), and multicollinearity (VIF,
when \(p>1\)) are checked as assumptions.

```{r}
test_linear_regression(sbp_3m ~ age + ldl, data = cardio)
```

## Logistic regression

`test_logistic_regression()` fits:

\[
\log\frac{P(y_i=1)}{1-P(y_i=1)}=\beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}
\]

with the overall likelihood-ratio test against the intercept-only model as
the primary result:

\[
LR = D_{null}-D_{model} \sim \chi^2_p, \qquad D=-2\log L
\]

\[
R^2_{McFadden}=1-\frac{\log L_{model}}{\log L_{null}}
\]

Coefficients are reported on both the log-odds and odds-ratio
(\(OR_j=e^{\beta_j}\)) scales. Multicollinearity (VIF) and influential
observations (Cook's distance, flagged at \(>4/n\)) are checked as
assumptions.

```{r}
test_logistic_regression(controlled_3m ~ age + ldl, data = cardio)
```

# Survival analysis

`make_cardio_data()` has no time-to-event columns, so this section uses a
small synthetic example. `Surv()` is re-exported from the `survival`
package.

```{r}
set.seed(1)
n <- 100
survival_dat <- tibble::tibble(
  time = rexp(n, 0.1),
  status = rbinom(n, 1, 0.7),
  arm = rep(c("control", "treatment"), each = n / 2),
  age = rnorm(n, 60, 10)
)
```

## Log-rank test

`test_survival()` compares two Kaplan-Meier survival curves via the
log-rank test:

\[
\chi^2 = \sum_{g\in\{A,B\}}\frac{(O_g-E_g)^2}{V_g} \sim \chi^2_1
\qquad\text{(Mantel, 1966)}
\]

with a companion univariate Cox hazard ratio \(HR=e^{\hat\beta}\) reported
as the effect size; the hazard ratio is not itself part of the log-rank
test and can disagree with it under strongly non-proportional hazards.

```{r}
test_survival(Surv(time, status) ~ arm, data = survival_dat)
```

## Cox proportional hazards regression

`test_cox()` leaves the baseline hazard \(h_0(t)\) unspecified:

\[
h(t\mid x) = h_0(t)\exp(\beta_1x_1+\cdots+\beta_px_p)
\]

Coefficients are reported as hazard ratios \(HR_j=e^{\beta_j}\); the primary
result is the overall likelihood-ratio test on \(p\) degrees of freedom.
Effect size is Harrell's concordance index. The proportional-hazards
assumption is checked via the correlation between scaled Schoenfeld
residuals and time (Grambsch & Therneau, 1994).

```{r}
test_cox(Surv(time, status) ~ age + arm, data = survival_dat)
```

# Diagnostic test accuracy and agreement

`make_cardio_data()` has no diagnostic-test or multi-rater columns, so this
section uses small synthetic examples.

## Diagnostic accuracy

With true/false positives/negatives \(TP,FP,FN,TN\) from a 2x2 test-vs.
-reference table:

\[
Sensitivity=\frac{TP}{TP+FN}, \quad Specificity=\frac{TN}{TN+FP}, \quad
PPV=\frac{TP}{TP+FP}, \quad NPV=\frac{TN}{TN+FN}
\]

\[
LR^+=\frac{Sensitivity}{1-Specificity}, \qquad LR^-=\frac{1-Sensitivity}{Specificity}
\]

Sensitivity, specificity, PPV, NPV, and accuracy each get an exact
(Clopper-Pearson) confidence interval; likelihood ratios use the closed-form
log-scale interval of Simel, Samsa & Matchar (1991),
\(SE[\log LR^+]=\sqrt{1/TP-1/(TP+FN)+1/FP-1/(FP+TN)}\). `test_diagnostic()`
tests overall accuracy against the no-information rate (the larger
reference-class proportion) as the primary result, matching
`caret::confusionMatrix()`'s convention.

```{r}
set.seed(1)
diag_dat <- tibble::tibble(
  test = c(rep("positive", 55), rep("negative", 98)),
  reference = c(rep("positive", 45), rep("negative", 10), rep("positive", 8), rep("negative", 90))
)
test_diagnostic(diag_dat, test, reference)
```

## ROC / AUC

`test_roc()` computes the AUC from the Mann-Whitney U relationship (the
probability a random positive-class observation outranks a random
negative-class one):

\[
AUC = \frac{U}{n_1n_0}
\]

with the closed-form Hanley & McNeil (1982) standard error:

\[
SE(AUC)=\sqrt{\frac{AUC(1-AUC)+(n_1-1)(Q_1-AUC^2)+(n_0-1)(Q_2-AUC^2)}{n_1n_0}},
\quad Q_1=\frac{AUC}{2-AUC}, \ Q_2=\frac{2AUC^2}{1+AUC}
\]

The optimal threshold maximizes Youden's J (1950),
\(J=Sensitivity+Specificity-1\).

```{r}
roc_dat <- tibble::tibble(
  marker = c(rnorm(60, 2, 1), rnorm(50, 0, 1)),
  disease = c(rep("yes", 60), rep("no", 50))
)
test_roc(roc_dat, marker, disease)
```

## Inter-rater agreement (Cohen's kappa)

`test_agreement()` reports Cohen's kappa from observed agreement \(p_o\)
and chance-expected agreement \(p_e\):

\[
\kappa = \frac{p_o-p_e}{1-p_e}
\]

Two *different* standard errors are used for two different purposes: the
confidence interval uses the large-sample SE of Fleiss, Cohen & Everitt
(1969), evaluated at \(\hat\kappa\); the z-test of \(H_0:\kappa=0\) uses the
smaller-magnitude null-hypothesis SE of Fleiss (1981), evaluated at
\(\kappa=0\). Reusing the CI SE for the null test (an error `testflow`
previously made and has since fixed) understates the null SE whenever
agreement is materially above chance, inflating the z-statistic and
anti-conservatively shrinking the p-value.

```{r}
agree_dat <- tibble::tibble(
  rater1 = sample(c("mild", "moderate", "severe"), 100, replace = TRUE),
  rater2 = sample(c("mild", "moderate", "severe"), 100, replace = TRUE)
)
test_agreement(agree_dat, rater1, rater2)
```

## Intraclass correlation coefficient

`test_icc()` reports three ICC forms from one-way and two-way ANOVA
variance decompositions (Shrout & Fleiss, 1979) of \(n\) subjects by \(k\)
raters:

\[
ICC(1,1)=\frac{BMS-WMS}{BMS+(k-1)WMS}
\]

\[
ICC(2,1)=\frac{BMS-EMS}{BMS+(k-1)EMS+\frac{k}{n}(JMS-EMS)}
\qquad\text{(two-way random, absolute agreement)}
\]

\[
ICC(3,1)=\frac{BMS-EMS}{BMS+(k-1)EMS}
\qquad\text{(two-way fixed, consistency)}
\]

ICC(2,1) is reported as the primary/effect-size estimate, following Koo &
Li's (2016) recommendation for typical reliability studies. Confidence
intervals use the F-distribution formulas of McGraw & Wong (1996), with a
Satterthwaite-approximated denominator degrees of freedom for ICC(2,1).

```{r}
icc_dat <- tibble::tibble(
  rater1 = rnorm(30, 50, 10),
  rater2 = rnorm(30, 50, 10),
  rater3 = rnorm(30, 50, 10)
)
test_icc(icc_dat, c(rater1, rater2, rater3))
```

# Outlier screening

`test_outliers()` is a screening workflow, not a hypothesis test: it flags
rows via the IQR rule (univariate, per variable),

\[
x < Q_1-1.5\,IQR \quad\text{or}\quad x > Q_3+1.5\,IQR
\]

and/or Mahalanobis distance (multivariate, `method = "mahalanobis"`),
flagged against a \(\chi^2\) cutoff at the 97.5th percentile with degrees of
freedom equal to the number of variables:

\[
D_i^2 = (x_i-\bar x)^\top S^{-1}(x_i-\bar x), \qquad
\text{flag if } D_i^2 > \chi^2_{0.975,\,p}
\]

```{r}
test_outliers(c(sbp_3m, ldl, crp), data = cardio)
```

# Effect-size reference

The table below consolidates every effect-size formula reported across the
workflows above, since some functions (Kendall's W, discordant pairs,
observed proportion) are only summarized inline elsewhere.

| Workflow | Effect size | Formula |
|---|---|---|
| One sample | Cohen's d | \(d=(\bar x-\mu_0)/s_x\) |
| Two groups (t-test) | Cohen's d | \(d=(\bar x_1-\bar x_2)/s_p\) |
| Two groups (Wilcoxon) | Rank-biserial r | \(r_{rb}=1-2W/(n_1n_2)\) |
| Paired | Cohen's dz | \(d_z=\bar d/s_d\) |
| ANOVA (one-way, factorial) | Eta squared | \(\eta_j^2=SS_j/\sum SS\) |
| Kruskal-Wallis | Epsilon squared | \(\epsilon^2=(H-k+1)/(n-k)\) |
| Repeated ANOVA | Eta squared (time) | \(\eta^2_{time}=SS_{time}/SS_{total}\) |
| Friedman | Kendall's W | \(W=\chi_F^2/(n(k-1))\) |
| Cochran Q | Kendall's-W analogue | \(W=Q/(n(k-1))\) |
| Categorical association | Cramer's V | \(V=\sqrt{\chi^2/(n(\min(r,c)-1))}\) |
| One-sample proportion | Observed proportion | \(\hat p=x/n\) |
| Multinomial | Chi-square GOF statistic | \(\chi^2=\sum_i(O_i-E_i)^2/E_i\) |
| Paired categorical | Discordant pairs | \(b+c\) |
| Correlation | r / rho / tau | see "Correlation" above |
| Linear regression | R squared, adjusted R squared | \(R^2=1-SS_{res}/SS_{total}\) |
| Logistic regression | McFadden's pseudo R squared | \(1-\log L_{model}/\log L_{null}\) |
| Survival (log-rank) | Hazard ratio | \(HR=e^{\hat\beta}\) (no magnitude label) |
| Cox regression | Concordance index | Harrell's C (no magnitude label) |
| Diagnostic accuracy | Accuracy | \((TP+TN)/n\) (no magnitude label) |
| ROC | AUC | \(U/(n_1n_0)\) |
| Agreement | Cohen's kappa | \((p_o-p_e)/(1-p_e)\) |
| ICC | ICC(2,1) | see "Intraclass correlation" above |

## Magnitude labels

- Cohen's \(d\): negligible \(<0.2\), small \(<0.5\), moderate \(<0.8\),
  otherwise large.
- Eta/epsilon/R-squared family: negligible \(<0.01\), small \(<0.06\),
  moderate \(<0.14\), otherwise large. (McFadden's pseudo \(R^2\) is not on
  the OLS \(R^2\) scale; values above ~0.2-0.4 are already a good fit in
  practice.)
- Cramer's V, rank-biserial \(|r|\), Kendall's W: negligible \(<0.1\), small
  \(<0.3\), moderate \(<0.5\), otherwise large.
- AUC: poor if \(\max(AUC,1-AUC)<0.7\), acceptable \(<0.8\), excellent
  \(<0.9\), otherwise outstanding (Hosmer, Lemeshow & Sturdivant 2013, p.
  177).
- Cohen's kappa: poor \(<0\), slight \(<0.21\), fair \(<0.41\), moderate
  \(<0.61\), substantial \(<0.81\), otherwise almost perfect (Landis & Koch,
  1977).
- ICC: poor \(<0.5\), moderate \(<0.75\), good \(<0.9\), otherwise excellent
  (Koo & Li, 2016).
- Hazard ratio, concordance index, accuracy: no magnitude label - there is
  no widely agreed small/moderate/large convention, and inventing one would
  overstate how standardized that judgment is. The estimate itself is
  always reported.

# Sample-size planning

`sample_size()` and the endpoint-specific `sample_size_*()` functions plan
continuous, binary, survival, ordinal, bioequivalence, and precision-based
designs, each under superiority, non-inferiority, or equivalence objectives
where applicable. Every result is a `sample_size` object with a raw and
dropout-adjusted \(n\), the assumptions used, a report-ready sentence, and
(for some designs) a power curve.

Notation: \(\alpha\) is the type I error rate, \(1-\beta\) is power,
\(z_{1-\alpha}\)/\(z_{1-\alpha/2}\)/\(z_{1-\beta}\) are standard normal
quantiles, \(r=n_B/n_A\) is the allocation ratio. Superiority uses two-sided
\(\alpha\); non-inferiority/equivalence use one-sided \(\alpha\) (pass
`alpha = 0.025` explicitly - `testflow` does not rescale it). Dropout
adjustment is always:

\[
n_{recruit} = \left\lceil \frac{n_{eval}}{1-d} \right\rceil
\]

```{r, eval = FALSE}
sample_size(
  endpoint = c("continuous", "binary", "survival", "ordinal"),
  design = c("parallel", "paired", "repeated"),
  objective = c("superiority", "noninferiority", "equivalence"),
  ...
)
```

## Continuous endpoints

`sample_size_continuous()` covers parallel, paired, and repeated designs.
For a parallel-group superiority comparison with common variance
\(\sigma^2\) and target difference \(\Delta\):

\[
n_A = \frac{(r+1)\sigma^2(z_{1-\beta}+z_{1-\alpha/2})^2}{r\Delta^2},
\qquad n_B = rn_A
\]

```{r}
sample_size_continuous(design = "parallel", objective = "superiority", delta = 5, sd = 10, alpha = 0.05, power = 0.90)
```

Paired designs replace \(\sigma\) with the paired-difference SD and drop the
allocation ratio:

\[
n_{pairs} = \frac{(z_{1-\alpha/2}+z_{1-\beta})^2\sigma_{diff}^2}{\Delta^2}
\]

For `design = "repeated"` with 3+ time points, \(\sigma_{diff}\) is rescaled
under a compound-symmetry working correlation before the paired formula is
applied (`testflow` design-effect extension, not a standalone textbook
formula):

\[
\sigma_{eff} = \sigma_{diff}\sqrt{\frac{1+(m-1)\rho}{m}}
\]

Non-inferiority uses \(d_{NI}=\Delta+\delta\); equivalence uses
\(d_{EQ}=\delta-|\Delta|\) (or, when \(\Delta=0\) exactly,
\(z_{1-\beta/2}\) in place of \(z_{1-\beta}\)), both plugged into the same
form as superiority with a one-sided \(\alpha\).

## Binary endpoints

`sample_size_binary()` covers parallel risk-difference and paired
(discordant-pairs) designs. Parallel superiority (`method = "pooled"`, with
\(\bar p=(p_A+rp_B)/(1+r)\)):

\[
n_A = \frac{\left[z_{1-\alpha/2}\sqrt{(1+1/r)\bar p(1-\bar p)}
+z_{1-\beta}\sqrt{p_A(1-p_A)+p_B(1-p_B)/r}\right]^2}{(p_A-p_B)^2}
\]

```{r}
sample_size_binary(design = "parallel", objective = "superiority", p1 = 0.4, p2 = 0.25, method = "pooled", alpha = 0.05, power = 0.90)
```

Paired binary designs plan on the discordant cells directly, with
discordant odds ratio \(OR_D=p_{10}/p_{01}\) and discordance rate
\(\lambda_D=p_{10}+p_{01}\):

\[
n_{total} = \left\lceil\frac{(z_{1-\alpha/2}+z_{1-\beta})^2(OR_D+1)^2}{(OR_D-1)^2\lambda_D}\right\rceil
\]

Non-inferiority/equivalence use the same asymmetric variance form at the
non-inferiority/equivalence distance instead of \(p_A-p_B\).

## Survival endpoints

`sample_size_survival()` plans required events, then converts to total N.
Under an assumed exponential hazard:

\[
D_{total} = \frac{4(z_{1-\alpha/2}+z_{1-\beta})^2}{[\log(HR)]^2}
\]

or, proportional-hazards-only (`method = "ph_only"`):

\[
D_{total} = \frac{2(HR+1)^2(z_{1-\alpha/2}+z_{1-\beta})^2}{(HR-1)^2}
\]

Given planned survival probabilities \(S_A(t)\), \(S_B(t)\), equal
allocation:

\[
N_{total} = \frac{2D_{total}}{(1-S_A(t))+(1-S_B(t))}
\]

```{r}
sample_size_survival(hr = 0.7, survival_a = 0.8, survival_b = 0.7, alpha = 0.05, power = 0.90)
```

With staggered uniform accrual (`accrual_duration`, `follow_up`), the flat
\(1-S_g(t)\) is replaced by the accrual-averaged event probability:

\[
\bar q_g = 1-\frac{e^{-\lambda_g(T-R)}-e^{-\lambda_gT}}{\lambda_gR}, \qquad T=R+\text{follow\_up}
\]

which always requires more subjects than instantaneous accrual for the same
event target, and reduces to the flat conversion as \(R\to0\).
Non-inferiority/equivalence work on the log-hazard-ratio scale with
\(d_{NI}=\log(HR_M)-\log(HR)\) or \(d_{EQ}=\min(\theta-\theta_L,\theta_U-\theta)\).

## Ordinal endpoints

`sample_size_ordinal()` implements Noether's method for a two-group
Mann-Whitney/Wilcoxon comparison, with \(P=P(A>B)+\tfrac12P(A=B)\):

\[
n_{per\ group} = \frac{(z_{1-\alpha/2}+z_{1-\beta})^2}{6(P-0.5)^2}
\]

```{r}
sample_size_ordinal(p_superiority = 0.65, alpha = 0.05, power = 0.90)
```

## Bioequivalence

`sample_size_bioequivalence()` plans a two one-sided test (TOST) study on
the log scale, with \(\theta_0=\log(GMR)\), bounds
\(\theta_L=\log(L)\)/\(\theta_U=\log(U)\) (usually 0.80/1.25), and
\(d_{BE}=\min(\theta_0-\theta_L,\theta_U-\theta_0)\). The default
`method = "iterative_tost"` searches for the smallest \(n\) achieving the
*exact* TOST power via the noncentral t distribution (Phillips 1990):

\[
Power(n) = T_{df,\delta_U}(-t_{1-\alpha,df}) - T_{df,\delta_L}(t_{1-\alpha,df})
\]

where \(T_{df,\delta}\) is the noncentral t CDF,
\(\delta_U=(\theta_0-\theta_U)/SE\), \(\delta_L=(\theta_0-\theta_L)/SE\),
\(\sigma=\sqrt{\log(1+CV^2)}\). Verified in package tests to match
`PowerTOST::power.TOST()` (the regulatory-standard, Owen's-Q-based
calculation) for balanced designs; `method = "normal_approx"` gives the
closed-form approximation \(n_{total}\approx2\sigma_w^2(z_{1-\beta}+z_{1-\alpha})^2/d_{BE}^2\),
which agrees closely near \(GMR=1\) but can diverge further away from it (an
earlier version of `testflow` used only the normal approximation here,
under-sizing studies by up to ~20% at small \(n\)).

```{r}
sample_size_bioequivalence(design = "crossover", gmr = 0.95, cv_within = 0.30, alpha = 0.05, power = 0.90)
```

## Precision-based sample size

`sample_size_precision()` sizes a study to a target confidence-interval
half-width \(w\) rather than power against an effect - there is no `power`
argument.

```{r}
sample_size_precision(endpoint = "continuous", design = "one_sample", width = 2, sd = 10, alpha = 0.05)
sample_size_precision(endpoint = "binary", design = "two_sample", width = 0.08, p1 = 0.4, p2 = 0.3, allocation = 2)
```

For `endpoint = "binary"`, `design = "one_sample"`, `method` chooses the
interval used for planning: `"wald"` (default, closed-form normal
approximation) or `"wilson"`/`"exact"` (Clopper-Pearson), which run a
deterministic search for the minimum integer \(n\) whose interval has
maximum one-sided half-width \(\leq w\) - important for rare-prevalence
studies (newborn screening, rare adverse events) where the Wald
approximation is unreliable:

```{r}
sample_size_precision(endpoint = "binary", design = "one_sample", width = 0.02, p = 0.01, method = "wilson")
```

`criterion = "expected"` (default) checks precision at the anticipated
event count `round(n*p)`; `criterion = "worst_case"` additionally requires
the target to hold across nearby plausible event counts. `x$diagnostics`
reports rare-event planning quantities a CI-width target alone does not
guarantee (`expected_events`, `probability_zero_events`, `complete_case_n`
vs. `adjusted_n`), with an `approximation_warning` when the Wald
approximation is used or expected counts are small.

For binary `design = "two_sample"`, with \(r=n_B/n_A\):

\[
n_A = \frac{z_{1-\alpha/2}^2}{w^2}\left[p_1(1-p_1)+\frac{p_2(1-p_2)}{r}\right],
\qquad n_B = rn_A
\]

## Cluster-randomized adjustment and dropout

`sample_size_cluster_adjust()` inflates an individually randomized \(n\) by
the design effect \(DE=1+(m-1)\rho\) (equal cluster size \(m\), intracluster
correlation \(\rho\)), or \(DE\approx1+((1+CV_m^2)m-1)\rho\) for unevenly
sized clusters. `sample_size_adjust_dropout()` exposes the dropout formula
above directly. Both return a plain integer, not a `sample_size` object -
they are post-processing adjustments, not full planning calculations.

```{r}
sample_size_cluster_adjust(100, m = 20, rho = 0.02)
sample_size_adjust_dropout(100, dropout = 0.15)
```

## Summary of supported settings

| Endpoint | Design(s) | Objective(s) | Formula family |
|---|---|---|---|
| Continuous | parallel, paired, repeated (2+ time points) | superiority, non-inferiority, equivalence | normal-approximation mean-difference formulas, compound-symmetry repeated extension |
| Binary | parallel, paired, repeated (2 time points) | superiority, non-inferiority, equivalence | risk-difference and discordant-pair planning |
| Survival | parallel | superiority, non-inferiority, equivalence | event-based proportional-hazards planning, optional uniform-accrual adjustment |
| Ordinal | parallel | superiority | Noether approximation |
| Bioequivalence | crossover, parallel | equivalence (TOST) | iterative TOST or normal approximation, log scale |
| Precision | one-sample, two-sample, log-OR | CI half-width (no power/objective) | continuous and binary CI-width formulas |

Every `sample_size` object also has `print()`, `summary()`, `report()`,
`as_tibble()`, and `plot(type = "summary"/"curve"/"both")` methods. See
`?sample_size_continuous`, `?sample_size_binary`, `?sample_size_survival`,
`?sample_size_ordinal`, `?sample_size_bioequivalence`,
`?sample_size_precision`, and `?sample_size_cluster_adjust` for the complete
argument reference and every `design`/`objective`/`method` combination.

# References

Gosset WS (Student). The probable error of a mean. *Biometrika*.
1908;6(1):1-25.

Welch BL. The generalization of "Student's" problem when several different
population variances are involved. *Biometrika*. 1947;34(1-2):28-35.

Wilcoxon F. Individual comparisons by ranking methods. *Biometrics
Bulletin*. 1945;1(6):80-83.

Mann HB, Whitney DR. On a test of whether one of two random variables is
stochastically larger than the other. *Annals of Mathematical Statistics*.
1947;18(1):50-60.

Fisher RA. *Statistical Methods for Research Workers*. Oliver and Boyd;
1925.

Levene H. Robust tests for equality of variances. In: *Contributions to
Probability and Statistics*. Stanford University Press; 1960.

Kruskal WH, Wallis WA. Use of ranks in one-criterion variance analysis.
*Journal of the American Statistical Association*. 1952;47(260):583-621.

Tukey JW. Comparing individual means in the analysis of variance.
*Biometrics*. 1949;5(2):99-114.

Dunn OJ. Multiple comparisons using rank sums. *Technometrics*.
1964;6(3):241-252.

Friedman M. The use of ranks to avoid the assumption of normality implicit
in the analysis of variance. *Journal of the American Statistical
Association*. 1937;32(200):675-701.

Girden ER. *ANOVA: Repeated Measures*. Sage; 1992.

Cochran WG. The comparison of percentages in matched samples. *Biometrika*.
1950;37(3/4):256-266.

McNemar Q. Note on the sampling error of the difference between correlated
proportions or percentages. *Psychometrika*. 1947;12(2):153-157.

Pearson K. Notes on regression and inheritance in the case of two parents.
*Proceedings of the Royal Society of London*. 1895;58:240-242.

Pearson K. On the criterion that a given system of deviations from the
probable in the case of a correlated system of variables is such that it
can be reasonably supposed to have arisen from random sampling.
*Philosophical Magazine*. 1900;50(302):157-175.

Fisher RA. On the interpretation of chi-square from contingency tables, and
the calculation of P. *Journal of the Royal Statistical Society*.
1922;85(1):87-94.

Spearman C. The proof and measurement of association between two things.
*The American Journal of Psychology*. 1904;15(1):72-101.

Kendall MG. A new measure of rank correlation. *Biometrika*.
1938;30(1/2):81-93.

Cramer H. *Mathematical Methods of Statistics*. Princeton University Press;
1946.

Clopper CJ, Pearson ES. The use of confidence or fiducial limits illustrated
in the case of the binomial. *Biometrika*. 1934;26(4):404-413.

Cohen J. *Statistical Power Analysis for the Behavioral Sciences* (2nd ed.).
Lawrence Erlbaum; 1988.

Draper NR, Smith H. *Applied Regression Analysis* (3rd ed.). Wiley; 1998.

Hosmer DW, Lemeshow S, Sturdivant RX. *Applied Logistic Regression* (3rd
ed.). Wiley; 2013.

Kaplan EL, Meier P. Nonparametric estimation from incomplete observations.
*Journal of the American Statistical Association*. 1958;53(282):457-481.

Mantel N. Evaluation of survival data and two new rank order statistics
arising in its consideration. *Cancer Chemotherapy Reports*.
1966;50(3):163-170.

Cox DR. Regression models and life-tables. *Journal of the Royal
Statistical Society: Series B*. 1972;34(2):187-202.

Grambsch PM, Therneau TM. Proportional hazards tests and diagnostics based
on weighted residuals. *Biometrika*. 1994;81(3):515-526.

Simel DL, Samsa GP, Matchar DB. Likelihood ratios with confidence: sample
size estimation for diagnostic test studies. *Journal of Clinical
Epidemiology*. 1991;44(8):763-770.

Altman DG, Bland JM. Diagnostic tests 1: sensitivity and specificity. *BMJ*.
1994;308(6943):1552.

Hanley JA, McNeil BJ. The meaning and use of the area under a receiver
operating characteristic (ROC) curve. *Radiology*. 1982;143(1):29-36.

Youden WJ. Index for rating diagnostic tests. *Cancer*. 1950;3(1):32-35.

Cohen J. A coefficient of agreement for nominal scales. *Educational and
Psychological Measurement*. 1960;20(1):37-46.

Fleiss JL, Cohen J, Everitt BS. Large sample standard errors of kappa and
weighted kappa. *Psychological Bulletin*. 1969;72(5):323-327.

Fleiss JL. *Statistical Methods for Rates and Proportions* (2nd ed.). Wiley;
1981.

Landis JR, Koch GG. The measurement of observer agreement for categorical
data. *Biometrics*. 1977;33(1):159-174.

Shrout PE, Fleiss JL. Intraclass correlations: uses in assessing rater
reliability. *Psychological Bulletin*. 1979;86(2):420-428.

McGraw KO, Wong SP. Forming inferences about some intraclass correlation
coefficients. *Psychological Methods*. 1996;1(1):30-46.

Koo TK, Li MY. A guideline of selecting and reporting intraclass
correlation coefficients for reliability research. *Journal of Chiropractic
Medicine*. 2016;15(2):155-163.

Julious SA. *Sample Sizes for Clinical Trials*. Chapman & Hall/CRC; 2010.

Phillips KF. Power of the two one-sided tests procedure in bioequivalence.
*Journal of Pharmacokinetics and Biopharmaceutics*. 1990;18(2):137-144.

Donner A, Klar N. *Design and Analysis of Cluster Randomization Trials in
Health Research*. Arnold; 2000.
